---
title: "Build and Deploy a Stroke Prediction Model Using R"
author: "Rohit Kotamsetti"
date: "`r Sys.Date()`"
output: html_document
---

# Data Analysis Report

This RMarkdown file documents the data analysis and steps taken to build and deploy a stroke prediction model using R. It includes tasks like data exploration, preprocessing, model building, evaluation, and deployment. The final report was completed on `r date()`.

### Data Description

According to the World Health Organization (WHO), stroke is the second leading cause of death globally, responsible for approximately 11% of total deaths.

This dataset aims to predict the likelihood of a patient experiencing a stroke based on various features such as gender, age, health conditions, and lifestyle factors like smoking status. Each row in the dataset contains relevant information about a patient.

---

# Task One: Import Data and Preprocessing

## Install and Load Required Packages

```{r}
# List of required packages
packages <- c("tidyverse", "lubridate", "ggplot2", "dplyr", "caret", "randomForest", "skimr", "gridExtra", "caTools", "corrplot", "ggcorrplot", "naniar")

# Install and load packages if not already installed
for (pkg in packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}

# Load the stroke dataset
stroke_data <- read_csv("healthcare-dataset-stroke-data.csv", show_col_types = FALSE)

# Preview the first few rows of the dataset
head(stroke_data)
```

## Describe and explore the data

```{r}
# Overview of the dataset
summary(stroke_data)  # Provides summary statistics
glimpse(stroke_data)  # Quick structure of the dataset
skim(stroke_data)     # Detailed summary of data (including missing values)

# Check for missing values in the dataset
miss_scan_count(data = stroke_data, search = list("Unknown", "N/A", "Other"))

# Convert 'bmi' to numeric and impute missing BMI values with the median
stroke_data$bmi <- as.numeric(stroke_data$bmi)
idx <- complete.cases(stroke_data)
bmi_idx <- is.na(stroke_data$bmi)
median_bmi <- median(stroke_data$bmi, na.rm = TRUE)
stroke_data[bmi_idx,]$bmi <- median_bmi

# Verify no missing values left in the dataset
colSums(is.na(stroke_data))

# Gender vs Stroke
stroke_data %>%
  ggplot(aes(x = gender, fill = factor(stroke))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("maroon", "blue")) +
  ggtitle("Gender vs. Stroke")

# Heart disease vs Stroke
stroke_data %>%
  ggplot(aes(x = factor(heart_disease), fill = factor(stroke))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("maroon", "blue")) +
  ggtitle("Heart Disease vs. Stroke")

# Residence Type vs Stroke
stroke_data %>%
  ggplot(aes(x = Residence_type, fill = factor(stroke))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("maroon", "blue")) +
  ggtitle("Residence Type vs. Stroke")

# Smoking Status vs Stroke
stroke_data %>%
  ggplot(aes(x = smoking_status, fill = factor(stroke))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("maroon", "blue")) +
  ggtitle("Smoking Status vs. Stroke")

# Average Glucose Level vs Stroke
stroke_data %>%
  ggplot(aes(x = avg_glucose_level, fill = factor(stroke))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("maroon", "blue")) +
  ggtitle("Average Glucose Level vs. Stroke")

# Body Mass Index vs Stroke (Filtered for valid BMI range)
stroke_data %>%
  filter(between(bmi, 0, 60)) %>%
  ggplot(aes(x = bmi, fill = factor(stroke))) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("maroon", "blue")) +
  ggtitle("Body Mass Index vs. Stroke")

# Remove 'id' column as it is not useful for prediction
stroke_data <- stroke_data %>%
  select(-c(id)) %>%
  filter(gender != "Other")  # Remove 'Other' category in gender column

# Convert categorical columns to factors
stroke_data$stroke <- factor(stroke_data$stroke, levels = c(0, 1), labels = c("No", "Yes"))
stroke_data$hypertension <- factor(stroke_data$hypertension, levels = c(0, 1), labels = c("No", "Yes"))
stroke_data$heart_disease <- factor(stroke_data$heart_disease, levels = c(0, 1), labels = c("No", "Yes"))

# Check the structure of the cleaned data
str(stroke_data)
```

# Task Two: Build prediction models

```{r}
# Distribution of Gender
d1 <- stroke_data %>%
  ggplot(aes(x = gender, fill = gender)) +
  geom_bar(fill = c("red", "blue")) +
  ggtitle("Gender Distribution") +
  geom_text(aes(label=..count..), stat = "Count", vjust = 1.0)

# Distribution of Hypertension
d2 <- stroke_data %>%
  ggplot(aes(x = hypertension, fill = hypertension)) +
  geom_bar(fill = c("red", "blue")) +
  ggtitle("Hypertension Distribution") +
  geom_text(aes(label=..count..), stat = "Count", vjust = 1.0)

# Display multiple plots in a grid
grid.arrange(d1, d2, ncol = 2)
```

# Task Three: Evaluate and select prediction models

```{r}
# Split the dataset into training (80%) and testing (20%) sets
sample.split(stroke_data$stroke, SplitRatio = 0.8) -> split_tag
train <- subset(stroke_data, split_tag == TRUE)
test <- subset(stroke_data, split_tag == FALSE)

# Check the dimensions of the train and test sets
dim(train)
dim(test)
```

# Task Four: Deploy the prediction model

```{r}
# Set seed for reproducibility
set.seed(123)

# Train a random forest model
rf <- randomForest(formula = stroke ~ ., data = train)

# Display the model summary
rf
```

# Task Five: Findings and Conclusions

```{r}
# Evaluate model performance on test data
confusionMatrix(predict(rf, test), test$stroke)
```